---
title: "DS 2020 Final Project Report"
author: "David Wang, Logan Pappas"
date: "2025-11-30"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load the data and library
library(tidyverse)

electric_vehicle_population_data <- read_csv("Electric_Vehicle_Population_Data.csv")

head(electric_vehicle_population_data)
glimpse(electric_vehicle_population_data)
dim(electric_vehicle_population_data)
```
##Introduction

Data is from https://catalog.data.gov/dataset/electric-vehicle-population-data 

This dataset shows the Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) that are currently registered through Washington State Department of Licensing (DOL). The vehicles can come from areas outside of the state.
Electric Range is the estimated distance a vehicle can travel on a single full charge powered solely by its battery before needing to be recharged (we measured by miles)
BEVs run purely on electricity, PHEVs are plug in hybrid electric vehicles that use both gas and electricity


## Topic: What are the differences in electric range across various vehicle makes, models, and location, along with other potential factors?

We want to know how these factors affect the electric range of a vehicle
Our data set includes these variables that we used to solve this question:
Make, Model, Electric Range, Electric Vehicle Type (BEV or PHEV), Model Year, City, County, Base MSRP

## Research Questions

* How does electric range differ across vehicle makes?
* How does electric range differ across models within the same make? 
* Do newer models within the same make tend to have higher range? 
* How does electric range differ between BEVs and PHEVs across makes? 
* Do average electric ranges differ across rural and urban cities?

```{r}
###cleaning to find median
electric_vehicle_population_data <- electric_vehicle_population_data %>%
  mutate(
    `Electric Range` = replace_na(`Electric Range`, median(`Electric Range`, na.rm = TRUE)),
    `Base MSRP` = replace_na(`Base MSRP`, median(`Base MSRP`, na.rm = TRUE))
  )

```

```{r}
#fixing data types
electric_vehicle_population_data <- electric_vehicle_population_data |>
  mutate(
    `Model Year` = as.integer(`Model Year`),
    `Postal Code` = as.character(`Postal Code`),
    `Legislative District` = as.factor(`Legislative District`),
    State = as.factor(State),
    `Electric Vehicle Type` = as.factor(`Electric Vehicle Type`),
    `Clean Alternative Fuel Vehicle (CAFV) Eligibility` =
      as.factor(`Clean Alternative Fuel Vehicle (CAFV) Eligibility`)
  )

#remove dupes
electric_vehicle_population_data <- electric_vehicle_population_data |>
  distinct(`VIN (1-10)`, `Model Year`, .keep_all = TRUE)

glimpse(electric_vehicle_population_data)

dim(electric_vehicle_population_data)
```
## How we Cleaned the Data/First Steps

1. Removed missing values in key numeric fields- Replaced missing Electric Range and Base MSRP values with the median (most were either 0 or NA)

2.Fixed data types – Converted year, postal code, and categorical fields so calculations and grouping work correctly.

3.Removed potential duplicate vehicles – Used VIN + Model Year so that each EV appears only once in the dataset.

4.Filtered invalid ranges – Kept only vehicles with electric range greater than 0 to focus on real EV performance data.

```{r}
library(tidyverse)
#we filter for vehicles that have electric range > 0 as we only care about electric cars
ev_plot <- electric_vehicle_population_data |>
  filter(!is.na(`Electric Range`), `Electric Range` > 0)

```

```{r}
#How does electric range differ across vehicle makes?

ev_plot |>
  ggplot(aes(x = reorder(Make, `Electric Range`, FUN = median),
             y = `Electric Range`)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Electric Range Distribution by Vehicle Make",
    x = "Make",
    y = "Electric Range (miles)"
  )

ev_plot |>
  group_by(Make) |>
  summarise(
    mean_range = mean(`Electric Range`),
    sd_range = sd(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(Make, mean_range), y = mean_range)) +
  geom_point(size = 2) +
  geom_errorbar(aes(
    ymin = mean_range - sd_range,
    ymax = mean_range + sd_range
  ), width = 0.2) +
  coord_flip() +
  labs(
    title = "Average Electric Range by Make (±1 SD)",
    x = "Make",
    y = "Average Electric Range (miles)"
  )


```

```{r}
#How does electric range differ across models within the same make?
top_makes <- ev_plot |>
  count(Make, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(Make)

ev_plot |>
  filter(Make %in% top_makes) |>
  ggplot(aes(
    x = `Electric Range`,
    y = interaction(Make, Model)
  )) +
  geom_boxplot() +
  facet_wrap(~ Make, scales = "free_y") +
  labs(
    title = "Electric Range by Model Within Top Vehicle Makes",
    x = "Electric Range (miles)",
    y = "Model"
  )

ev_plot |>
  filter(Make %in% top_makes) |>
  group_by(Make, Model) |>
  summarise(
    avg_range = mean(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = avg_range, y = Model)) +
  geom_point() +
  facet_wrap(~ Make, scales = "free_y") +
  labs(
    title = "Average Electric Range by Model Within Each Make",
    x = "Average Electric Range (miles)",
    y = "Model"
  )


```

```{r}
#Do average electric ranges differ across rural and urban cities?
urban_rural_lookup <- tibble(
  City = c("Seattle", "Tacoma", "Spokane", "Bellevue", "Olympia",
           "Orondo", "Orondo", "Orting", "Palouse", "Port Hadlock"),
  Urban_Rural = c("Urban", "Urban", "Urban", "Urban", "Urban",
                  "Rural", "Rural", "Rural", "Rural", "Rural")
)
ev_city_classified <- electric_vehicle_population_data %>%
  left_join(urban_rural_lookup, by = "City")

ev_city_classified %>%
  filter(!is.na(Urban_Rural)) %>%
  group_by(Urban_Rural) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ggplot(aes(x = Urban_Rural, y = avg_range, fill = Urban_Rural)) +
  geom_col() +
  labs(
    title = "Average Electric Range: Urban vs Rural Cities",
    x = "City Type",
    y = "Average Range (miles)"
  ) +
  scale_fill_manual(values = c("Urban" = "steelblue", "Rural" = "forestgreen"))

ev_city_classified %>%
  filter(!is.na(Urban_Rural)) %>%
  group_by(City, Urban_Rural) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(City, avg_range), y = avg_range, fill = Urban_Rural)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Electric Range of Selected Urban and Rural Cities",
    x = "City",
    y = "Average Electric Range"
  )
electric_vehicle_population_data %>%
  count(County) %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%   # top 20 counties
  ggplot(aes(x = reorder(County, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 Counties by Number of Registered EVs",
    x = "County",
    y = "Number of EVs"
  ) +
  theme_minimal()
```
1Rural cities show a higher average electric range compared to urban cities in our subset.
-Rural areas averaged about 49 miles, while urban areas averaged around 34 miles.
1This may suggest that drivers in rural areas tend to own EVs with longer ranges, possibly due to greater travel distances or fewer charging stations 
while people in more urban areas may not need to 
travel as far to get to their destination and have 
access to more charging stations.


```{r}
#Do newer models within the same make tend to have higher range?

ev_plot |>
  filter(Make %in% top_makes) |>
  group_by(Make, `Model Year`) |>
  summarise(
    avg_range = mean(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = `Model Year`, y = avg_range, color = Make)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Electric Range Over Time by Make",
    x = "Model Year",
    y = "Average Electric Range (miles)"
  )

```

```{r}
#How does electric range differ between BEVs and PHEVs across makes?
ev2 <- electric_vehicle_population_data %>%
  mutate(EV_Type = case_when(
    `Electric Vehicle Type` == "Battery Electric Vehicle (BEV)" ~ "BEV",
    `Electric Vehicle Type` == "Plug-in Hybrid Electric Vehicle (PHEV)" ~ "PHEV",
    TRUE ~ NA_character_
  ))

avg_range_by_type <- ev2 %>%
  filter(!is.na(EV_Type)) %>%
  group_by(EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE))

p1 <- ggplot(avg_range_by_type,
             aes(x = EV_Type, y = avg_range, fill = EV_Type)) +
  geom_col() +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  labs(
    title = "Average Electric Range: BEVs vs PHEVs",
    x = "Vehicle Type",
    y = "Average Electric Range (miles)"
  ) +
  theme_minimal()

top10_BEV <- ev2 %>%
  filter(EV_Type == "BEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

top10_PHEV <- ev2 %>%
  filter(EV_Type == "PHEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

range_top20 <- ev2 %>%
  filter(
    (EV_Type == "BEV" & Make %in% top10_BEV) |
    (EV_Type == "PHEV" & Make %in% top10_PHEV)
  ) %>%
  group_by(Make, EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ungroup()

unique_levels <- unique(c(top10_BEV, top10_PHEV))

range_top20$Make <- factor(range_top20$Make, levels = unique_levels)

p2 <- ggplot(range_top20,
             aes(x = Make, y = avg_range, fill = EV_Type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Top 10 BEV vs Top 10 PHEV Makes — Average Electric Range",
    x = "Vehicle Make",
    y = "Average Electric Range (miles)",
    fill = "EV Type"
  ) +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  theme_minimal()

p1
p2
```
-Battery Electric Vehicles (BEVs) have a noticeably higher average range than Plug-in Hybrid Electric Vehicles (PHEVs).
-In our dataset, BEVs average around 38 miles of electric range, while PHEVs average closer to 30 miles.
-This aligns with expectations: BEVs rely fully on electric power, so manufacturers maximize their battery capacity, while PHEVs use smaller batteries since they 
also have gasoline engines.

