---
title: "DS 2020 Final Project Report"
author: "David Wang, Logan Pappas"
date: "2025-11-30"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# load the data and library
library(tidyverse)

electric_vehicle_population_data <- read_csv("Electric_Vehicle_Population_Data.csv")

head(electric_vehicle_population_data)
glimpse(electric_vehicle_population_data)
dim(electric_vehicle_population_data)
```
## Introduction

Data is from https://catalog.data.gov/dataset/electric-vehicle-population-data 

This dataset shows the Battery Electric Vehicles (BEVs) and Plug-in Hybrid Electric Vehicles (PHEVs) that are currently registered through Washington State Department of Licensing (DOL). The vehicles can come from areas outside of the state.
Electric Range is the estimated distance a vehicle can travel on a single full charge powered solely by its battery before needing to be recharged (we measured by miles)
BEVs run purely on electricity, PHEVs are plug in hybrid electric vehicles that use both gas and electricity


## Topic: What are the differences in electric range across various vehicle makes, models, and location, along with other potential factors?

We want to know how these factors affect the electric range of a vehicle
Our data set includes these variables that we used to solve this question:
Make, Model, Electric Range, Electric Vehicle Type (BEV or PHEV), Model Year, City, County, Base MSRP

## Research Questions

* How does electric range differ across vehicle makes?
* How does electric range differ across models within the same make? 
* Do newer models within the same make tend to have higher range? 
* How does electric range differ between BEVs and PHEVs across makes? 
* Do average electric ranges differ across rural and urban cities?

```{r}
###cleaning to find median
electric_vehicle_population_data <- electric_vehicle_population_data %>%
  mutate(
    `Electric Range` = replace_na(`Electric Range`, median(`Electric Range`, na.rm = TRUE)),
    `Base MSRP` = replace_na(`Base MSRP`, median(`Base MSRP`, na.rm = TRUE))
  )

```

```{r}
#fixing data types
electric_vehicle_population_data <- electric_vehicle_population_data |>
  mutate(
    `Model Year` = as.integer(`Model Year`),
    `Postal Code` = as.character(`Postal Code`),
    `Legislative District` = as.factor(`Legislative District`),
    State = as.factor(State),
    `Electric Vehicle Type` = as.factor(`Electric Vehicle Type`),
    `Clean Alternative Fuel Vehicle (CAFV) Eligibility` =
      as.factor(`Clean Alternative Fuel Vehicle (CAFV) Eligibility`)
  )

#remove dupes
electric_vehicle_population_data <- electric_vehicle_population_data |>
  distinct(`VIN (1-10)`, `Model Year`, .keep_all = TRUE)

glimpse(electric_vehicle_population_data)

dim(electric_vehicle_population_data)
```
## How we Cleaned the Data/First Steps

1. Removed missing values in key numeric fields- Replaced missing Electric Range and Base MSRP values with the median (most were either 0 or NA)

2.Fixed data types – Converted year, postal code, and categorical fields so calculations and grouping work correctly.

3.Removed potential duplicate vehicles – Used VIN + Model Year so that each EV appears only once in the dataset.

4.Filtered invalid ranges – Kept only vehicles with electric range greater than 0 to focus on real EV performance data.

```{r}
library(tidyverse)
#we filter for vehicles that have electric range > 0 as we only care about electric cars
ev_plot <- electric_vehicle_population_data |>
  filter(!is.na(`Electric Range`), `Electric Range` > 0)

```

```{r}
#How does electric range differ across vehicle makes?

ev_plot |>
  ggplot(aes(x = reorder(Make, `Electric Range`, FUN = median),
             y = `Electric Range`)) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Electric Range Distribution by Vehicle Make",
    x = "Make",
    y = "Electric Range (miles)"
  )

ev_plot |>
  group_by(Make) |>
  summarise(
    mean_range = mean(`Electric Range`),
    sd_range = sd(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = reorder(Make, mean_range), y = mean_range)) +
  geom_point(size = 2) +
  geom_errorbar(aes(
    ymin = mean_range - sd_range,
    ymax = mean_range + sd_range
  ), width = 0.2) +
  coord_flip() +
  labs(
    title = "Average Electric Range by Make (±1 SD)",
    x = "Make",
    y = "Average Electric Range (miles)"
  )

#filter top 15 median electric range vehicles

top_15_makes <- ev_plot |>
  group_by(Make) |>
  summarise(
    median_range = median(`Electric Range`),
    .groups = "drop"
  ) |>
  arrange(desc(median_range)) |>
  slice_head(n = 15) |>
  pull(Make)

ev_plot |>
  filter(Make %in% top_15_makes) |>
  ggplot(aes(
    x = reorder(Make, `Electric Range`, FUN = median),
    y = `Electric Range`
  )) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Top 15 Vehicle Makes by Median Electric Range",
    x = "Make",
    y = "Electric Range (miles)"
  )

ev_plot |>
  filter(Make %in% top_15_makes) |>
  group_by(Make) |>
  summarise(
    mean_range = mean(`Electric Range`),
    sd_range = sd(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(
    x = reorder(Make, mean_range),
    y = mean_range
  )) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_range - sd_range,
      ymax = mean_range + sd_range
    ),
    width = 0.2
  ) +
  coord_flip() +
  labs(
    title = "Top 15 Vehicle Makes by Average Electric Range (±1 SD)",
    x = "Make",
    y = "Average Electric Range (miles)"
  )


#decided to just leave in popular makes, which tend to have the highest variablility + to make graph bigger and more visible

# identify the top 15 vehicle makes by IQR (boxplot spread)
top_15_iqr_makes <- ev_plot |>
  group_by(Make) |>
  summarise(
    q1 = quantile(`Electric Range`, 0.25),
    q3 = quantile(`Electric Range`, 0.75),
    iqr = q3 - q1,
    .groups = "drop"
  ) |>
  arrange(desc(iqr)) |>
  slice_head(n = 15) |>
  pull(Make)

# boxplot for these top 15 makes by IQR
ev_plot |>
  filter(Make %in% top_15_iqr_makes) |>
  ggplot(aes(
    x = reorder(
      Make,
      `Electric Range`,
      FUN = function(x) quantile(x, 0.75) - quantile(x, 0.25)
    ),
    y = `Electric Range`
  )) +
  geom_boxplot() +
  coord_flip() +
  labs(
    title = "Vehicle Makes with the Largest Electric Range Variability",
    x = "Make",
    y = "Electric Range (miles)"
  )


```

## Graph Findings (based off Vehicle Makes with Largest Electric Range Variability)

* Filtered down to makes that have the highest variability, which also tend to be in more popular makes

* Electric range varies based on manufacturers, makes like Tesla, Volkswagen, Nissan have higher median electric range

* In the same make, electric range can range quite a bit, with Audi and Hyundai have the widest range of  electric range

* These results suggest that electric range depends not just on manufacturer, but also on specific models and vehicle type.

* Because of outliers and uneven distributions of how models are offered, comparisons between manufacturers should be interpreted cautiously.


```{r}
#How does electric range differ across models within the same make?
top_makes <- ev_plot |>
  count(Make, sort = TRUE) |>
  slice_head(n = 6) |>
  pull(Make)

ev_plot |>
  filter(Make %in% top_makes) |>
  ggplot(aes(
    x = `Electric Range`,
    y = interaction(Make, Model)
  )) +
  geom_boxplot() +
  facet_wrap(~ Make, scales = "free_y") +
  labs(
    title = "Electric Range by Model Within Top Vehicle Makes",
    x = "Electric Range (miles)",
    y = "Model"
  )

ev_plot |>
  filter(Make %in% top_makes) |>
  group_by(Make, Model) |>
  summarise(
    avg_range = mean(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = avg_range, y = Model)) +
  geom_point() +
  facet_wrap(~ Make, scales = "free_y") +
  labs(
    title = "Average Electric Range by Model Within Each Make",
    x = "Average Electric Range (miles)",
    y = "Model"
  )


```

## Graph Findings

* Electric range can vary a lot across models with the same manufacturer, showing that not all models that have the same make perform similarly.

* Manufacturers like Tesla and Chevrolet, show clear electric range differences between models.

* Other makes, like Audi, BMW, and Volvo, tend to be at low electric ranges, maybe due to being plug-in hybrids or shorter-range models.

* A lot of makes have certain models that stand out as high-range outliers, which can disproportionately influence overall summaries.


```{r}
#Do average electric ranges differ across rural and urban cities?
urban_rural_lookup <- tibble(
  City = c("Seattle", "Tacoma", "Spokane", "Bellevue", "Olympia",
           "Orondo", "Orondo", "Orting", "Palouse", "Port Hadlock"),
  Urban_Rural = c("Urban", "Urban", "Urban", "Urban", "Urban",
                  "Rural", "Rural", "Rural", "Rural", "Rural")
)
ev_city_classified <- electric_vehicle_population_data %>%
  left_join(urban_rural_lookup, by = "City")

ev_city_classified %>%
  filter(!is.na(Urban_Rural)) %>%
  group_by(Urban_Rural) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ggplot(aes(x = Urban_Rural, y = avg_range, fill = Urban_Rural)) +
  geom_col() +
  labs(
    title = "Average Electric Range: Urban vs Rural Cities",
    x = "City Type",
    y = "Average Range (miles)"
  ) +
  scale_fill_manual(values = c("Urban" = "steelblue", "Rural" = "forestgreen"))

ev_city_classified %>%
  filter(!is.na(Urban_Rural)) %>%
  group_by(City, Urban_Rural) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ggplot(aes(x = reorder(City, avg_range), y = avg_range, fill = Urban_Rural)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Average Electric Range of Selected Urban and Rural Cities",
    x = "City",
    y = "Average Electric Range"
  )
electric_vehicle_population_data %>%
  count(County) %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%   # top 20 counties
  ggplot(aes(x = reorder(County, n), y = n)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Top 20 Counties by Number of Registered EVs",
    x = "County",
    y = "Number of EVs"
  ) +
  theme_minimal()
```

## Graph Findings
* Rural cities show a higher average electric range compared to urban cities in our subset.
* Rural areas averaged about 49 miles, while urban areas averaged around 34 miles.
* This may suggest that drivers in rural areas tend to own EVs with longer ranges, possibly due to greater travel distances or fewer charging stations 
while people in more urban areas may not need to 
travel as far to get to their destination and have 
access to more charging stations.


```{r}
#Do newer models within the same make tend to have higher range?

ev_plot |>
  filter(Make %in% top_makes) |>
  group_by(Make, `Model Year`) |>
  summarise(
    avg_range = mean(`Electric Range`),
    .groups = "drop"
  ) |>
  ggplot(aes(x = `Model Year`, y = avg_range, color = Make)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Electric Range Over Time by Make",
    x = "Model Year",
    y = "Average Electric Range (miles)"
  )

```

## Graph Findings

* Average electric range tends to increase over time for most manufacturers, which reflects improvements in battery technology. 

* Tesla consistently shows higher average ranges than other makes, while makes with sharp increases can be due to the company investing more into creating electric vehicles

* Overall, trends vary widely across manufacturers, so year-to-year changes should be interpreted cautiously.


```{r}
#How does electric range differ between BEVs and PHEVs across makes?
ev2 <- electric_vehicle_population_data %>%
  mutate(EV_Type = case_when(
    `Electric Vehicle Type` == "Battery Electric Vehicle (BEV)" ~ "BEV",
    `Electric Vehicle Type` == "Plug-in Hybrid Electric Vehicle (PHEV)" ~ "PHEV",
    TRUE ~ NA_character_
  ))


avg_range_by_type <- ev2 %>%
  filter(!is.na(EV_Type)) %>%
  group_by(EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE))

p1 <- ggplot(avg_range_by_type,
             aes(x = EV_Type, y = avg_range, fill = EV_Type)) +
  geom_col() +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  labs(
    title = "Average Electric Range: BEVs vs PHEVs",
    x = "Vehicle Type",
    y = "Average Electric Range (miles)"
  ) +
  theme_minimal()

top10_BEV <- ev2 %>%
  filter(EV_Type == "BEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

top10_PHEV <- ev2 %>%
  filter(EV_Type == "PHEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

range_top20 <- ev2 %>%
  filter(
    (EV_Type == "BEV" & Make %in% top10_BEV) |
    (EV_Type == "PHEV" & Make %in% top10_PHEV)
  ) %>%
  group_by(Make, EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ungroup()

unique_levels <- unique(c(top10_BEV, top10_PHEV))

range_top20$Make <- factor(range_top20$Make, levels = unique_levels)

p2 <- ggplot(range_top20,
             aes(x = Make, y = avg_range, fill = EV_Type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Top 10 BEV vs Top 10 PHEV Makes — Average Electric Range",
    x = "Vehicle Make",
    y = "Average Electric Range (miles)",
    fill = "EV Type"
  ) +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  theme_minimal()

p1
p2
ev2 <- electric_vehicle_population_data %>%
  mutate(EV_Type = case_when(
    `Electric Vehicle Type` == "Battery Electric Vehicle (BEV)" ~ "BEV",
    `Electric Vehicle Type` == "Plug-in Hybrid Electric Vehicle (PHEV)" ~ "PHEV",
    TRUE ~ NA_character_
  ))

# NEW LINE: remove Tesla for the fair comparison
ev_no_tesla <- ev2 %>% filter(Make != "TESLA")

# ---- Original BEV vs PHEV graph (WITH Tesla) ----
avg_range_by_type <- ev2 %>%
  filter(!is.na(EV_Type)) %>%
  group_by(EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE))

p1 <- ggplot(avg_range_by_type,
             aes(x = EV_Type, y = avg_range, fill = EV_Type)) +
  geom_col() +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  labs(
    title = "Average Electric Range: BEVs vs PHEVs",
    x = "Vehicle Type",
    y = "Average Electric Range (miles)"
  ) +
  theme_minimal()


# ---- Original top 10 BEV / PHEV makes (WITH Tesla) ----
top10_BEV <- ev2 %>%
  filter(EV_Type == "BEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

top10_PHEV <- ev2 %>%
  filter(EV_Type == "PHEV") %>%
  count(Make, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(Make)

range_top20 <- ev2 %>%
  filter(
    (EV_Type == "BEV" & Make %in% top10_BEV) |
    (EV_Type == "PHEV" & Make %in% top10_PHEV)
  ) %>%
  group_by(Make, EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE)) %>%
  ungroup()

unique_levels <- unique(c(top10_BEV, top10_PHEV))

range_top20$Make <- factor(range_top20$Make, levels = unique_levels)

p2 <- ggplot(range_top20,
             aes(x = Make, y = avg_range, fill = EV_Type)) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Top 10 BEV vs Top 10 PHEV Makes — Average Electric Range",
    x = "Vehicle Make",
    y = "Average Electric Range (miles)",
    fill = "EV Type"
  ) +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  theme_minimal()


# ---- NEW GRAPH: BEVs vs PHEVs with Tesla REMOVED ----
avg_range_no_tesla <- ev_no_tesla %>%
  group_by(EV_Type) %>%
  summarise(avg_range = mean(`Electric Range`, na.rm = TRUE))

ggplot(avg_range_no_tesla,
       aes(x = EV_Type, y = avg_range, fill = EV_Type)) +
  geom_col() +
  scale_fill_manual(values = c("BEV" = "steelblue", "PHEV" = "darkorange")) +
  labs(
    title = "Average Electric Range: BEVs vs PHEVs (Tesla Removed)",
    x = "Vehicle Type",
    y = "Average Electric Range (miles)"
  ) +
  theme_minimal()

# Show original graphs
p1
p2
```

## Graph Findings
* Battery Electric Vehicles (BEVs) have a noticeably higher average range than Plug-in Hybrid Electric Vehicles (PHEVs).
* In our dataset, BEVs average around 38 miles of electric range, while PHEVs average closer to 30 miles.
* This aligns with expectations: BEVs rely fully on electric power, so manufacturers maximize their battery capacity, while PHEVs use smaller batteries since they 
also have gasoline engines.

## Shortcomings/Outliers
* The MSRP variable had almost no data making it hard to connect any variables to price
* Tesla often threw averages off having a much higher range than any other brand
* Having many models and cities in the data set, we often had to restructure our graphs to make it readable
* Only showing top models isn’t necessarily representative of all the models present
* The dataset we chose is only for registered cars in Washington so it isn’t a good representation of all cars in the US

## Conclusion
Overall, electric range varies significantly by make, model, vehicle type, and location. We found that

* The make influences the range
* Models from the same make influences the range
* The average range for each model tends to increase over time
* BEVs have a higher average range than PHEV’s
* Rural cities show higher average range than Urban cities

## Further Questions and Research
* How does the price affect the range?
* How could charging station availability affect the range?
* Could weather or temperature affect the range?
* How do these cars compare to gas powered cars?
* What areas/counties have the most electric vehicles?


